\documentclass{ximera}

\input{../../preamble.tex}


\author{Bart Snapp}



\title{Vector spaces}


\begin{document}
\begin{abstract}
  We describe properties of vectors and matrices and discuss their
  implications.
\end{abstract}
\maketitle

\begin{quote}
  Think deeply about simple things.

  \hfill ---A.E.\ Ross

\end{quote}


Data and their associated systems of equations are complicated. It is
not enough to simply solve these equations, we must understand the
solutions in a more abstract way. In this chapter we will introduce
various sets (or in this case, vector spaces) of vectors in the
context of solving equations. Then we will revist these ideas from
`higher ground,' and introduce the notion of a vector space.



\section{Null space}

Suppose we have a system of equations:
\begin{align*}
  a_1 x + b_1 y + c_1 z &= d_1\\
  a_2 x + b_2 y + c_2 z &= d_2\\
  a_3 x + b_3 y + c_3 z &= d_3
\end{align*}
Where each $a_i$, $b_i$, $c_i$, and $d_i$ are real numbers. We can
learn about the solutions to this system of equations if we
\textit{change the problem} to the following:
\begin{align*}
  a_1 x + b_1 y + c_1 z &= 0  \\
  a_2 x + b_2 y + c_2 z &= 0  \\
  a_3 x + b_3 y + c_3 z &= 0
\end{align*}



\section{Column space}

Given a system of equations
\begin{align*}
  a_1 x + b_1 y + c_1 z &= d_1\\
  a_2 x + b_2 y + c_2 z &= d_2\\
  a_3 x + b_3 y + c_3 z &= d_3
\end{align*}
the \dfn{column space} is the set of all possible constants, expressed
as a single vector:
\[
\begin{pmatrix} d_1 \\ d_2 \\ d_3 \end{pmatrix}
\]



\section{Row space}




\section{Linear combinatrions of vectors}

\begin{definition}\index{K-vector space@$K$-vector space}\index{vector space@$K$-vector space}
  A \textbf{$\boldsymbol{K}$-vector space} is an Abelian group $(V,+)$
  with identity $\vec{0}$, along with a field $K$ such that we may
  multiply group elements by field elements, meaning that there is a
  binary operation $-\cdot-: K\times V \to V$ such that if $\nu,\mu\in
  V$ and $a,b,\in K$ we have:
\begin{description}
\item[Compatibility with scalars] $(ab)\cdot \nu = a\cdot (b\cdot \nu)$.
\item[Vectors distribute over scalars] $(a+b)\cdot \nu =
  a\cdot\nu + b\cdot \nu$.
\item[Scalars distribute over vectors] $a\cdot (\nu+\mu) =
  a\cdot \nu + a\cdot \mu$.
\item[Identity is respected] $1_K\cdot \nu = \nu$.
\end{description}
In this case, elements of the group $V$ are called \dfn{vectors} and
elements of the field $K$ are called \dfn{scalars}.
\end{definition}

\begin{lemma}[Subspace criterion]\index{subspace criterion}
  Let $V$ be a $K$-vector space. $W\subset V$ is a subspace of $V$ if
  and only if
  \begin{enumerate}
  \item $W\ne \emptyset$.
  \item $W$ is closed under multiplication by scalars.
  \item $W$ is closed under vector addition.
  \end{enumerate}
\end{lemma}

\subsection{Span of vectors}

\begin{definition}
  Given a set of vectors $S$, in a $K$-vector space, $V$, the
  \dfn{span} of the vectors in $S$ is
  \[
  \Span(S) = \left\{\sum_{i=1}^n a_i\sigma_i:\text{$n\in \N$,
    $\sigma_i\in S$, and $a_i\in K$}\right\}.
  \]
  If $\Span(S) = V$, then we say $S$ is a \dfn{spanning set}.
\end{definition}

\subsection{Linear indendence of vectors}


\begin{definition}
  Given a $K$-vector space $V$, a finite set of vectors
  \[
  \{\lambda_1,\dots,\lambda_n\}
  \]
  is said to be \dfn{linearly independent} if
  \[
  a_1\lambda_1 + a_2\lambda_2 +\cdots + a_n\lambda_n = 0\quad \Rightarrow \quad a_1= \cdots =a_n = 0.
  \]
  A finite set of vectors is set to be \dfn{linearly dependent} if
  they are not linearly independent.
\end{definition}



\section{Column space and range of a matrix}

\subsection{Singular matrices}

\end{document}
